Efficient analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr

### 2018-01-29 cabios-template2/bigsnpr-review.tex
changes with Mike

{\color{red}
Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals.
As a consequence, it has been required to continuously optimize algorithms and this has led to the re-implementation of some software that would have become obsolete otherwise. For computing Principal Component Analysis (PCA), commonly performed to account for population stratification in association, a fast mode named FastPCA has been added to smartpca, and FlashPCA has been replaced by FlashPCA2 \cite[]{Abraham2014a,Abraham2016a,Galinsky2016,Price2006}. 
PLINK 1.07, which has been a central tool in the analysis of genotype data, has been replaced by PLINK 1.9, and soon PLINK 2.0 \cite[]{chang2015second,Purcell2007}.

Increasing size of genetic datasets is a source of major computational challenges and many analytical tools would be restricted by the amount of memory (RAM) available on computers.
This is particularly a burden for commonly used analysis languages such as R.
For analyzing genotype datasets in R, a range of software are available, including R packages GenABEL, SNPRelate and GWASTools \cite[]{aulchenko2007genabel,Gogarten2012,zheng2012high}. Package GenABEL, which has been widely used and cited, does not handle standard binary PLINK files and has not been updated since 2013.
Solving the memory issues for languages such as R would give access to a broad range of already implemented tools for data analysis. Fortunately, strategies have been developed to avoid loading large datasets in RAM. For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read or write operations. Storing large matrices on disk and accessing them via memory-mapping has been available for several years in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}.
}

@@

{\color{red}
In order to perform analyses of large-scale genomic data in R, we developed two R packages, bigstatsr and bigsnpr, that provide a wide-range of building blocks which are parts of standard analyses. 
For instance, one might like using R because this programming language makes it easy to tie together (existing or new) functions to be used as part of large, interactive and reproducible analyses.
We provide a similar format as filebacked ``big.matrix'' objects that we called ``Filebacked Big Matrices (FBMs)''. Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data. This data format is a particularly good trade-off between easiness of use and computation efficiency, making our code both simple and fast.
}
Package bigstatsr implements many statistical tools for several types of FBMs (unsigned char, unsigned short, integer and double). This includes implementation of multivariate sparse linear models, principal component analysis, association tests, matrix operations, and numerical summaries. The statistical tools developed in bigstatsr can be used for other types of data as long as they can be represented as matrices. Package bigsnpr depends on bigstatsr, using a special type of FBM object to store the genotypes, called ``FBM.code256''. Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (thinning, testing, predicting, plotting). 
We use both a real case-control genomic dataset for celiac disease and large-scale simulated data to illustrate application of the two R packages, including two association studies and the computation of Polygenic Risk Scores (PRS). We compare results from bigstatsr and bigsnpr with those obtained when using PLINK, EIGENSOFT and PRSice. We report execution times along with the code to perform major computational tasks.

@@

For preprocessing steps, we think the user should use PLINK, which is an invaluable tool for this. For the sake of reproducibility, one could use PLINK directly from R via systems calls. For the sake of simplicity, we provide wrappers as R functions that use system calls to PLINK for the conversion and quality control steps with as input a variety of formats (e.g.\ vcf, bed/bim/fam, ped/map) and bed/bim/fam files as output (Figure~\ref{fig:qc}).
Package bigsnpr provides fast conversions between bed/bim/fam PLINK files and the ``bigSNP'' object, which contains the genotype Filebacked Big Matrix (FBM.code256), a data frame with information on samples and another data frame with information on SNPs. We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``FBM''. Finally, we provide two methods for reading-in dosages as proof of concept and will provide a more final solution when we get access to the UK Biobank dataset ([SUPPMAT]).

Most modern SNP chips provide genotype data with large call-rates. For example, the Celiac data we use in this paper presents only a proportion of 0.0004 of missing values for the genotyped SNPs after quality control. Yet, most of the functions in bigstatsr and bigsnpr do not handle missing values.
So, we provide two functions for imputing missing values of genotyped SNPs. Note that we do not impute completely missing SNPs which would require the use of reference panels and could be performed via e.g.\ imputation servers for human data \cite[]{mccarthy2016reference}. The first function is a wrapper to PLINK and Beagle \cite[]{Browning2009} which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R (Figure~\ref{fig:impute}). Beagle uses phasing before imputing, which could take several weeks even when there are only a tiny amount of missing values to be imputed. The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy. We also provide an accurate estimation of the number of errors of imputation by SNP for post quality control.
This algorithm is based on Machine Learning approaches for genetic imputation \cite[]{Wang2012} and does not use phasing, thus allowing for a dramatic decrease in computation time. It only relies on some local XGBoost models. XGBoost is an optimized distributed gradient boosting library that can be used in R and provides some of the best results in machine learning competitions \cite[]{Chen2016}.
XGBoost builds decision trees that can detect nonlinear interactions, partially reconstructing phase, making it well suited for imputing genotype matrices. 
Our algorithm is the following: for each SNP, we divide the individuals in the ones which have a missing genotype (test set) and the ones which have a non-missing genotype for this particular SNP. Those latter individuals are further separated in a training set and a validation set (e.g.\ 80\% training and 20\% validation). The training set is used to build the XGBoost model for predicting missing data. The prediction model is then evaluated on the validation set for which we know the true genotype values, which gives an unbiased estimator of the number of genotypes that have been wrongly imputed for that particular SNP. The prediction model is also projected on the test set (missing values) in order to impute them.
}

@@

SNP thinning improves ascertainment of population structure with PCA \cite[]{Abdellaoui2013}. There are at least 3 different approaches to thin SNPs based on Linkage Disequilibrium. Two of them, named pruning and clumping, address SNPs in LD close to each others because of recombination events, while the third one address long-range regions with a complex LD pattern due to other biological events such as inversions \cite[]{Price2008}. 
First, pruning is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning based on a given threshold of correlation.
{\color{red}
Clumping is useful if a statistic is available for sorting the SNPs by importance. Clumping is usually used to post-process results of a GWAS in order to keep only the most significant SNP per region of the genome. 
For PCA, the thinning procedure should remain unsupervised (one must not use any phenotype) and we therefore propose to use the minor allele frequency (MAF) as the statistic of importance. 
Note that this choice does not come out of nowhere because in the pruning algorithm of PLINK, when two nearby SNPs are correlated, only the one with the highest MAF is kept.
Yet, in some worst-case scenario, the pruning algorithm can leave regions of the genome without any representative SNP at all ([SUPMAT]). 
So, we suggest to always use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping of package bigsnpr. In practice, we tested the effect of using clumping with the MAF instead of pruning for the thinning step before PCA and the results were very similar.
}

As mentioned above, the third approach that is generally combined with pruning and consists of removing SNPs in long-range LD regions \cite[]{Price2008}. Long-range LD regions for the human genome are available as an online table\footnote{https://goo.gl/8TngVE} that package bigsnpr can use to discard SNPs in these regions before computing PCs. 
However, the pattern of LD might be population specific, so we developed an iterative algorithm that automatically detects these long-range LD regions and removes them. This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping (with MAFs), then outliers SNPs are detected using the robust Mahalanobis distance as implemented in method pcadapt \cite[]{Luu2017}. Finally, the algorithm considers that consecutive outlier SNPs are in long-range LD regions. Indeed, a long-range LD region would cause SNPs in this region to have strong consecutive weights (loadings) in the PCA. This algorithm is implemented in function snp\_autoSVD of package bigsnpr and will be referred by this name in the rest of the paper.

@@

Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries. Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcomes \cite[]{Zheng2012}. 
Package bigstatsr implements statistical tests based on linear and logistic regressions. For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where
\begin{multline}
  \hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
\end{multline}
and $K$ is the number of principal components and $L$ is the number of other covariates (such as age and gender). Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where
\begin{multline}
  \log{\left(\frac{\hat{p}}{1-\hat{p}}\right)} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
\end{multline}
and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.
{\color{red}
These tests can be used to perform genome-wide association studies (GWAS) and are very fast due to the use of ingenious implementations, partly based on previous work by \cite{sikorska2013gwas}. 
}
%For a continuous phenotype $y$ and a SNP $s$ and some covariables $X$ (including a column of 1s for the intercept), we can compute $s^* = s - X(X^TX)^{-1} X^T s$ and $y^* = s - X(X^TX)^{-1} X^T y$ (orthogonal projections relative to the space of covariables).

@@

In this paper, two datasets are analyzed: the celiac disease cohort and the POPRES datasets \cite[]{Dubois2010,Nelson2008}. The Celiac dataset is composed of 15,283 individuals of European ancestry genotyped on 295,453 SNPs. The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.
For computation time comparisons, we replicated individuals in the Celiac dataset 5 and 10 times in order to increase sample size while keeping the same {\color{red} eigen decomposition (up to a constant) and pairwise SNP correlations} as the original dataset. To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also through replication of the Celiac dataset. \label{sec:rep}

@@

We present the results for four different analyses. 
First, we illustrate the application of R packages bigstatsr and bigsnpr. 
Secondly, by performing two GWAS, we compare the performance of bigstatsr and bigsnpr to the performance obtained with FastPCA (EIGENSOFT 6.1.4) and PLINK 1.9, and also two R packages SNPRelate and GWASTools \cite[]{chang2015second,Galinsky2016,Gogarten2012,zheng2012high}. For the step concerning the computation of 10 principal components, we further compare methods on larger datasets.
Thirdly, by performing a standard PRS analysis, we compare the performance of bigstatsr and bigsnpr to the performance obtained with PRSice-2 \cite[]{Euesden2015}. 
Finally, we present results of the two new methods implemented in bigsnpr, one method for the automatic detection and removal of long-range LD regions in PCA and another for the in-sample imputation of missing genotypes (i.e.\ for genotyped SNPs only). 
We use three types of data: a case-control cohort for the celiac disease, the European population cohort POPRES and simulated datasets using real genotypes from the Celiac cohort. 
We compare performance on two computers, a desktop computer with 64GB of RAM and 12 cores (6 physical cores), and a laptop with only 8GB of RAM and 4 cores (2 physical cores). For the functions that enable parallelism, we use half of the cores available on the corresponding computer.

@@

The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs with more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency (MAF) lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $10^{-10}$, and finally, removing the first individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}. 
For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\%. The 0.04\% missing genotype values were imputed with the XGBoost method. If we used a standard R matrix to store the genotypes, this data would require 32GB of memory. On the disk, the ``.bed'' file requires 1GB and the ``.bk'' file (storing the FBM) requires 4GB.

@@

%The Celiac dataset is relatively small as compared to modern genetic cohorts. 
To illustrate the scalability of the two R packages, we performed a GWAS analysis on 500K individuals and 1M SNPs. The GWAS analysis completed in approximately 11 hours using the aforementioned desktop computer. The GWAS analysis was composed of four main steps. 
First we convert binary PLINK files in our format ``bigSNP'' in 1 hour.
Then, we removed SNPs in long-range LD regions and used SNP clumping, leaving 93,083 SNPs in 5.4h. Then, the 10 first PCs were computed on the 500K individuals and these remaining SNPs in 1.8h. Finally, we performed a linear association test on the complete 500K dataset for each of the 1M SNPs, using the 10 first PCs as covariables in 2.9h.

@@

First, we compared the GWAS computations obtained with bigstatsr and bigsnpr to the ones obtained with PLINK 1.9 and EIGENSOFT 6.1.4, and also two R packages SNPRelate and GWASTools.
For most functions, multithreading is not available yet in PLINK, nevertheless, PLINK-specific algorithms that use bitwise parallelism (e.g.\ pruning) are still faster than the parallel algorithms reimplemented in package bigsnpr (Table~\ref{tab:bench-gwas}). 
Overall, performing a GWAS on a binary outcome with our two R packages, bigstatsr and bigsnpr, is as fast as when using EIGENSOFT and PLINK, and much faster than when using R packages SNPRelate and GWASTools. 
For performing an association study on a continuous outcome, we report a dramatic increase in performance by using bigstatsr and bigsnpr, making it possible to perform such analysis in less than 2 minutes for a relatively large dataset such as the Celiac dataset (Table~\ref{tab:bench-gwas}).
Note that the PC scores obtained are more precise as compared to PLINK (see the last paragraph of this subsection), which is also the case for the p-values computed for the two GWAS (see supplementary notebook ``GWAS-comparison'').

@@

Finally, on our desktop computer, we compared the computation times of FastPCA (fast mode of EIGENSOFT), FlashPCA2 and PLINK 2.0 (approx mode) to the similar function big\_randomSVD implemented in bigstatsr. For each comparison, we used the 93,083 SNPs which were remaining after pruning and we computed 10 PCs. We used the datasets of growing size simulated from the Celiac dataset (from 15,155 to 151,550 individuals). Overall, our function  big\_randomSVD showed to be almost twice as fast as FastPCA and FlashPCA2 and 8 times as fast when using parallelism with 6 cores, an option not currently possible with either FastPCA or FlashPCA2 (Figure~\ref{fig:bench-pca}). PLINK 2.0 proved to be faster than bigstatsr with a decrease in time of 20-40\%. 
We also compared results in terms of precision by comparing squared correlation between approximated PCs and ``true'' PCs provided by an exact eigen decomposition obtained with PLINK 2.0 (exact mode). 
Package bigstatsr and FlashPCA2 (that use the same algorithm) infer all PCs with a squared correlation of more than 0.999 between true PCs and approximated ones (Figure~\ref{fig:prec-pca}). Yet, FastPCA (fast mode of EIGENSOFT) and PLINK 2.0 (that use the same algorithm) infer the true first 6 PCs but the squared correlation between true PCs and approximated ones decreases for further PCs (Figure~\ref{fig:prec-pca}).

@@

For detecting long-range LD regions during the computation of PCA, we tested the function snp\_autoSVD on both the Celiac and POPRES datasets. For the POPRES dataset, the algorithm converged in two iterations. The first iterations found 3 long-range LD regions in chromosomes 2, 6 and 8 (Table~\ref{tab:lrldr-popres}). 
We compared the PCs of genotypes obtained after applying snp\_autoSVD with the PCs obtained after removing pre-determined long-range LD regions\footnote{https://goo.gl/8TngVE} and found a mean correlation of 89.6\% between PCs, mainly due to a rotation of PC7 and PC8 (Table~\ref{tab:pc-popres}). 
For the Celiac dataset, we found 5 long-range LD regions (Table~\ref{tab:lrldr-celiac}) and a mean correlation of 98.6\% between PCs obtained with snp\_autoSVD and the ones obtained by clumping and removing of predetermined long-range LD regions (Table~\ref{tab:pc-celiac}).

>>>>

{\color{red}
Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals.
As a consequence, it has been required to continuously optimize algorithms and this has led to the re-implementation of some software that would have become obsolete otherwise. For computing Principal Component Analysis (PCA), commonly performed to account for population stratification in association, a fast mode named FastPCA has been added to the software EIGENSOFT, and FlashPCA has been replaced by FlashPCA2 \cite[]{Abraham2014a,Abraham2016a,Galinsky2016,Price2006}. 
PLINK 1.07, which has been a central tool in the analysis of genotype data, has been replaced by PLINK 1.9 to speed-up computations, and there is also an alpha version of PLINK 2.0 that will handle more data types \cite[]{chang2015second,Purcell2007}.

Increasing size of genetic datasets is a source of major computational challenges and many analytical tools would be restricted by the amount of memory (RAM) available on computers.
This is particularly a burden for commonly used analysis languages such as R.
For analyzing genotype datasets in R, a range of software are available, including R packages GenABEL, SNPRelate and GWASTools \cite[]{aulchenko2007genabel,Gogarten2012,zheng2012high}. The popular package GenABEL does not handle standard binary PLINK files and has not been updated since 2013.
Solving memory issues for languages such as R would give access to a broad range of already implemented tools for data analysis. Fortunately, strategies have been developed to avoid loading large datasets in RAM. For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read or write operations. Storing large matrices on disk and accessing them via memory-mapping has been available for several years in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}.
}

@@

{\color{red}
In order to perform analyses of large-scale genomic data in R, we developed two R packages, bigstatsr and bigsnpr, that provide a wide-range of building blocks which are parts of standard analyses. 
R is a programming language that makes it easy to tie together existing or new functions to be used as part of large, interactive and reproducible analyses \cite[]{R2017}.
We provide a similar format as filebacked ``big.matrix'' objects that we called ``Filebacked Big Matrices (FBMs)''. Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data. This data format is a particularly good trade-off between easiness of use and computation efficiency, making our code both simple and fast.
}
Package bigstatsr implements many statistical tools for several types of FBMs (unsigned char, unsigned short, integer and double). This includes implementation of multivariate sparse linear models, principal component analysis, association tests, matrix operations, and numerical summaries. The statistical tools developed in bigstatsr can be used for other types of data as long as they can be represented as matrices. Package bigsnpr depends on bigstatsr, using a special type of FBM object to store the genotypes, called ``FBM.code256''. Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (thinning, testing, predicting, plotting). 
We use both a real case-control genomic dataset for celiac disease and large-scale simulated data to illustrate application of the two R packages, including two association studies and the computation of Polygenic Risk Scores (PRS). We compare results from bigstatsr and bigsnpr with those obtained when using PLINK, EIGENSOFT and PRSice. We report execution times along with the code to perform major computational tasks.

@@

For preprocessing steps, PLINK is a widely-used software. For the sake of reproducibility, one could use PLINK directly from R via systems calls. For the sake of simplicity, we provide wrappers as R functions that use system calls to PLINK for the conversion and quality control steps with as input a variety of formats (e.g.\ vcf, bed/bim/fam, ped/map) and bed/bim/fam files as output (Figure~\ref{fig:qc}).
Package bigsnpr provides fast conversions between bed/bim/fam PLINK files and the ``bigSNP'' object, which contains the genotype Filebacked Big Matrix (FBM.code256), a data frame with information on samples and another data frame with information on SNPs. We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``FBM''. Finally, we provide two methods for reading dosage data ([SUPPMAT]).

Most modern SNP chips provide genotype data with large call-rates. For example, the Celiac data we use in this paper presents only a proportion of 0.0004 of missing values for the genotyped SNPs after quality control. Yet, most of the functions in bigstatsr and bigsnpr do not handle missing values.
So, we provide two functions for imputing missing values of genotyped SNPs. Note that we do not impute completely missing SNPs which would require the use of reference panels and could be performed via e.g.\ imputation servers for human data \cite[]{mccarthy2016reference}. The first function is a wrapper to PLINK and Beagle \cite[]{Browning2009} which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R (Figure~\ref{fig:impute}). Beagle uses phasing before imputing, which could require several weeks of computation even when there are only a tiny amount of missing values to be imputed. The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy. We also provide an accurate estimation of the number of errors of imputation by SNP for post quality control.
This algorithm is based on Machine Learning approaches for genetic imputation \cite[]{Wang2012} and does not use phasing, thus allowing for a dramatic decrease in computation time. It only relies on some local XGBoost models. XGBoost is an optimized distributed gradient boosting library that can be used in R and provides some of the best results in machine learning competitions \cite[]{Chen2016}.
XGBoost builds decision trees that can detect nonlinear interactions, partially reconstructing phase, making it well suited for imputing genotype matrices. 
Our algorithm is the following: for each SNP, we divide the individuals in the ones which have a missing genotype (test set) and the ones which have a non-missing genotype for this particular SNP. Those latter individuals are further separated in a training set and a validation set (e.g.\ 80\% training and 20\% validation). The training set is used to build the XGBoost model for predicting missing data. The prediction model is then evaluated on the validation set for which we know the true genotype values, which gives an unbiased estimator of the number of genotypes that have been wrongly imputed for that particular SNP. The prediction model is also projected on the test set (missing values) in order to impute them.
}

@@

SNP thinning improves ascertainment of population structure with PCA \cite[]{Abdellaoui2013}. There are at least 3 different approaches to thin SNPs based on Linkage Disequilibrium. Two of them, named pruning and clumping, address SNPs in LD close to each others because of recombination events, while the third one address long-range regions with a complex LD pattern due to other biological events such as inversions \cite[]{Price2008}. 
First, pruning is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning based on a given threshold of correlation.
{\color{red}
Clumping is useful if a statistic is available for sorting the SNPs by importance. Clumping is usually used to post-process results of a GWAS in order to keep only the most significant SNP per region of the genome. 
For PCA, the thinning procedure should remain unsupervised (one must not use any phenotype) and we therefore propose to use the minor allele frequency (MAF) as the statistic of importance. 
This choice is consistent with the pruning algorithm of PLINK; when two nearby SNPs are correlated, PLINK keeps only the one with the highest MAF.
Yet, in some worst-case scenario, the pruning algorithm can leave regions of the genome without any representative SNP at all ([SUPMAT]). 
So, we suggest to always use clumping instead of pruning, using the MAF as the statistic of importance, which is the default in function snp\_clumping of package bigsnpr. In practice, for the three datasets we considered, the clumping algorithm with the MAF provides similar sets of SNPs as when using the pruning algorithm (result not shown).
}

The third approach, which is generally combined with pruning, consists of removing SNPs in long-range LD regions \cite[]{Price2008}. Long-range LD regions for the human genome are available as an online table\footnote{https://goo.gl/8TngVE} that package bigsnpr can use to discard SNPs in these regions before computing PCs. 
However, the pattern of LD might be population specific, so we developed an iterative algorithm that automatically detects these long-range LD regions and removes them. This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping (with MAFs), then outliers SNPs are detected using the robust Mahalanobis distance as implemented in method pcadapt \cite[]{Luu2017}. Finally, the algorithm considers that consecutive outlier SNPs are in long-range LD regions. Indeed, a long-range LD region would cause SNPs in this region to have strong consecutive weights (loadings) in the PCA. This algorithm is implemented in function snp\_autoSVD of package bigsnpr and will be referred by this name in the rest of the paper.

@@

Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries. Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcomes in bigsnpr \cite[]{Zheng2012}. 
Package bigstatsr implements statistical tests based on linear and logistic regressions. For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where
\begin{multline}
  \hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
\end{multline}
and $K$ is the number of principal components and $L$ is the number of other covariates (such as age and gender). Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where
\begin{multline}
  \log{\left(\frac{\hat{p}}{1-\hat{p}}\right)} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K \\ + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,
\end{multline}
and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.
{\color{red}
These tests can be used to perform genome-wide association studies (GWAS) and are very fast due to the use of ingenious implementations, partly based on previous work by \cite{sikorska2013gwas}. 
}
%For a continuous phenotype $y$ and a SNP $s$ and some covariables $X$ (including a column of 1s for the intercept), we can compute $s^* = s - X(X^TX)^{-1} X^T s$ and $y^* = s - X(X^TX)^{-1} X^T y$ (orthogonal projections relative to the space of covariables).

@@

In this paper, two datasets are analyzed: the celiac disease cohort and the POPRES datasets \cite[]{Dubois2010,Nelson2008}. The Celiac dataset is composed of 15,283 individuals of European ancestry genotyped on 295,453 SNPs. The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.
For computation time comparisons, we replicated individuals in the Celiac dataset 5 and 10 times in order to increase sample size while keeping the same {\color{red} eigen decomposition (up to a constant) and pairwise SNP correlations} as the original dataset. To assess scalibility of the packages for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also through replication of the Celiac dataset. \label{sec:rep}

@@

We present the results of four different analyses. 
First, we illustrate the application of R packages bigstatsr and bigsnpr. 
Secondly, by performing two GWAS, we compare the performance of bigstatsr and bigsnpr to the performance obtained with FastPCA (EIGENSOFT 6.1.4) and PLINK 1.9, and also two R packages SNPRelate and GWASTools \cite[]{chang2015second,Galinsky2016,Gogarten2012,zheng2012high}. PCA is a computationally intensive step of the GWAS, so that we further compare PCA methods on larger datasets.
Thirdly, by performing a standard PRS analysis, we compare the performance of bigstatsr and bigsnpr to the performance obtained with PRSice-2 \cite[]{Euesden2015}. 
Finally, we present results of the two new methods implemented in bigsnpr, one method for the automatic detection and removal of long-range LD regions in PCA and another for the in-sample imputation of missing genotypes (i.e.\ for genotyped SNPs only). 
We compare performance on two computers, a desktop computer with 64GB of RAM and 12 cores (6 physical cores), and a laptop with only 8GB of RAM and 4 cores (2 physical cores). For the functions that enable parallelism, we use half of the cores available on the corresponding computer.

@@

The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs with more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency (MAF) lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $10^{-10}$, and finally, removing the first individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}. 
For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\%. The 0.04\% missing genotype values were imputed with the XGBoost method. If we would have used a standard R matrix to store the genotypes, this data would have required 32GB of memory. On the disk, the ``.bed'' file requires 1GB and the ``.bk'' file (storing the FBM) requires 4GB.

@@

%The Celiac dataset is relatively small as compared to modern genetic cohorts. 
To illustrate the scalability of the two R packages, we performed a GWAS analysis on 500K individuals and 1M SNPs. The GWAS analysis completed in approximately 11 hours using the aforementioned desktop computer. The GWAS analysis was composed of four main steps. 
First we convert binary PLINK files in the format ``bigSNP'' in 1 hour.
Then, we removed SNPs in long-range LD regions and used SNP clumping, leaving 93,083 SNPs in 5.4h. Then, the 10 first PCs were computed on the 500K individuals and these remaining SNPs in 1.8h. Finally, we performed a linear association test on the complete 500K dataset for each of the 1M SNPs, using the 10 first PCs as covariables in 2.9h.

@@

First, we compared the GWAS computations obtained with bigstatsr and bigsnpr to the ones obtained with PLINK 1.9 and EIGENSOFT 6.1.4, and also two R packages SNPRelate and GWASTools.
For most functions, multithreading is not available yet in PLINK, nevertheless, PLINK-specific algorithms that use bitwise parallelism (e.g.\ pruning) are still faster than the parallel algorithms reimplemented in package bigsnpr (Table~\ref{tab:bench-gwas}). 
Overall, performing a GWAS on a binary outcome with bigstatsr and bigsnpr is as fast as when using EIGENSOFT and PLINK, and 19-45 times faster than when using R packages SNPRelate and GWASTools. 
For performing an association study on a continuous outcome, we report a dramatic increase in performance by using bigstatsr and bigsnpr, making it possible to perform such analysis in less than 2 minutes for a relatively large dataset such as the Celiac dataset. This analysis was 7-19 times faster as compared to PLINK 1.9 and 28-74 times faster as compared to SNPRelate and GWASTools (Table~\ref{tab:bench-gwas}).
Note that the PC scores obtained are more precise as compared to PLINK (see the last paragraph of this subsection), which is also the case for the p-values computed for the two GWAS (see supplementary notebook ``GWAS-comparison'').

@@

Finally, on our desktop computer, we compared the computation times of FastPCA (fast mode of EIGENSOFT), FlashPCA2 and PLINK 2.0 (approx mode) to the similar function big\_randomSVD implemented in bigstatsr. For each comparison, we used the 93,083 SNPs which were remaining after pruning and we computed 10 PCs. We used the datasets of growing size simulated from the Celiac dataset (from 15,155 to 151,550 individuals). Overall, function  big\_randomSVD is almost twice as fast as FastPCA and FlashPCA2 and 8 times as fast as when using parallelism with 6 cores, an option not currently available in either FastPCA or FlashPCA2 (Figure~\ref{fig:bench-pca}). PLINK 2.0 is faster than bigstatsr with a decrease in time of 20-40\%. 
We also compared results in terms of precision by comparing squared correlation between approximated PCs and ``true'' PCs provided by an exact eigen decomposition obtained with PLINK 2.0 (exact mode). 
Package bigstatsr and FlashPCA2 (that use the same algorithm) infer all PCs with a squared correlation of more than 0.999 between true PCs and approximated ones (Figure~\ref{fig:prec-pca}). Yet, FastPCA (fast mode of EIGENSOFT) and PLINK 2.0 (that use the same algorithm) infer the true first 6 PCs but the squared correlation between true PCs and approximated ones decreases for further PCs (Figure~\ref{fig:prec-pca}).

@@

For detecting long-range LD regions during the computation of PCA, we tested the function snp\_autoSVD on both the Celiac and POPRES datasets. For the POPRES dataset, the algorithm converged in two iterations. The first iterations found 3 long-range LD regions in chromosomes 2, 6 and 8 (Table~\ref{tab:lrldr-popres}). 
We compared the PCs of genotypes obtained after applying snp\_autoSVD with the PCs obtained after removing pre-determined long-range LD regions\footnote{https://goo.gl/8TngVE} and found a mean correlation of 89.6\% between PCs, mainly due to a rotation of PC7 and PC8 (Table~\ref{tab:pc-popres}). 
For the Celiac dataset, we found 5 long-range LD regions (Table~\ref{tab:lrldr-celiac}) and a mean correlation of 98.6\% between PCs obtained with snp\_autoSVD and the ones obtained by clumping and removing predetermined long-range LD regions (Table~\ref{tab:pc-celiac}).
